---
title: "Zanoni Graphs"
author: "James R. Hunter"
date: "September 12, 2016"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Modules
```{r packages}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(DescTools))
suppressPackageStartupMessages(library(ggbeeswarm))
```

## Load Data
```{r loaddata}
#, eval = FALSE}
mzdata <- read_csv("fc_final.csv")
# Correct gene names
mzdata$Symbol[mzdata$Symbol == "ELANE"] <- "ELA2"
mzdata$Symbol[mzdata$Symbol == "SERPINC1"] <- "ATIII"
mzdata$Symbol[mzdata$Symbol == "TNFSF10"] <- "TRAIL"
mzdata$Symbol[mzdata$Symbol == "TNFRSF1B"] <- "TNFR2"
# Re-save mzdata with corrections
save(mzdata, file = "mzdata.RData")
```

```{r loadcorrecteddata}
load("mzdata.RData")
kable(mzdata, caption = "Graph Data - Gene Expression")
```

## Difference in Expression between Elite Controllers (EC) and Rapid Progressors (RP)

```{r ecrp}
# Calculate difference using both recent and 12 month data
# Use absolute value of differences since direction doesn't count
mzdata <- mzdata %>% mutate(diffecrpri = abs(RPRI - ECRI)) %>%
                     mutate(diffecrp12 = abs(RP12 - EC12))
# Create frame with 20 largest differences for each period
# These need to be long data frames for the graphs
ecrpri <- top_n(mzdata, 20, diffecrpri) %>%
          select(Position, Symbol, ECRI, RPRI, diffecrpri) %>%
          gather(key = gr, value = exp, 3:4)
ecrp12 <- top_n(mzdata, 20, diffecrp12) %>%
          select(Position, Symbol, EC12, RP12, diffecrp12) %>%
          gather(key = gr, value = exp, 3:4)
```
```{r gr1}
# make bar graph showing difference for these twenty genes
# use data from 12 months
ecrp12gr <- ggplot(data = ecrp12, aes(x = Symbol, y = exp, fill = gr))
ecrp12gr <- ecrp12gr + scale_fill_brewer(palette = "Set1")
ecrp12gr <- ecrp12gr + geom_bar(stat = "identity", position = "dodge")
ecrp12gr <- ecrp12gr + xlab("20 Genes with Greatest Difference") + ylab("Expression")
ecrp12gr <- ecrp12gr + ggtitle("Difference in Expression between\nElite Controllers and Rapid Progressors - 12 months")
ecrp12gr <- ecrp12gr + labs(fill = "Group")
ecrp12gr <- ecrp12gr + theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
ecrp12gr
```
### Do the same for Recent Infection

```{r gr2}
# make bar graph showing difference for these twenty genes
# use data from 12 months
ecrprigr <- ggplot(data = ecrpri, aes(x = Symbol, y = exp, fill = gr))
ecrprigr <- ecrprigr + scale_fill_brewer(palette = "Set1")
ecrprigr <- ecrprigr + geom_bar(stat = "identity", position = "dodge")
ecrprigr <- ecrprigr + xlab("20 Genes with Greatest Difference") + ylab("Expression")
ecrprigr <- ecrprigr + ggtitle("Difference in Expression between\nElite Controllers and Rapid Progressors - Recent Infection")
ecrprigr <- ecrprigr + labs(fill = "Group")
ecrprigr <- ecrprigr + theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
ecrprigr
```

## Are the groups significantly different?

```{r ecrpdiff}
# make long version of ecrp for difference tests
ecrp_long <- mzdata %>% 
             select(Symbol, ECRI, EC12, RPRI, RP12) %>% 
             gather(key = gr, value = expr, 2:5) %>%
             mutate(term = substr(gr, 3, 4))
## 12 months
df <- ecrp_long %>% filter(term == "12")
(tt12 <- t.test(expr ~ gr, data = df))
## Recent Infection
df <- ecrp_long %>% filter(term == "RI")
(ttri <- t.test(expr ~ gr, data = df))

```
### Are the periods significantly different?

```{r ecrpterm}
(ttterm <- t.test(expr ~ term, data = ecrp_long))

```

## Determine Genes that Are Significantly Different between EC and RP
### Recent Infection
### DeltaDeltaCT (2^-DDCT)

```{r modecrpriddct}
# read file from Excel
ecrpriddct <- read_csv("ecrpriddct.csv")
genes <- ecrpriddct$Symbol
patients <- colnames(ecrpriddct[2:11])

# set up data frame of t-tests
ttdf <- data_frame(Symbol = genes, tstat = 0.0, pval = 0.0, lowerci = 0.0,
                   upperci = 0.0, sd1 = 0.0, sd2 = 0.0)

# transform matrix to put genes as variables and patients as observations
x <- as_data_frame(t(as.matrix(ecrpriddct[,2:11])))
colnames(x) <- genes

# set up to loop t-tests over genes
for (i in 1:length(genes)) {
  # set up data frame for this gene from transformed matrix
  # columns 1:5 are group 1 and cols 6:10 are group 2
  # get group 1 data
  ec <- x %>% slice(1:5) %>% select(i)
  ec <- unlist(ec)
  
  # get group 2 data
  rp <- x %>% slice(6:10) %>% select(i)
  rp <- unlist(rp)
  
  # Calculate standard deviations for both groups and place in data frame
  ttdf$sd1[i] <- sd(ec, na.rm = TRUE)
  ttdf$sd2[i] <- sd(rp, na.rm = TRUE)
  
  # run t.test and load values in data frame
  xt <- t.test(ec, rp)
  ttdf$tstat[i] <- xt$statistic
  ttdf$pval[i] <- xt$p.value
  ttdf$lowerci[i] <- xt$conf.int[1]
  ttdf$upperci[i] <- xt$conf.int[2]
}

```

## Identify genes that are statistically significant

```{r modsiggenes}
siggenes <- ttdf %>% 
            filter(pval < 0.05) %>%
            arrange(pval)
kable(siggenes, caption = "Statistics to Determine Significant Genes")
# Create data frame with expression data for these significant genes
genessig <- siggenes$Symbol
ecrprisig <- mzdata %>% filter(Symbol %in% genessig) %>%
             select(Position, Symbol, ECRI, RPRI, diffecrpri) %>%
             gather(key = gr, value = exp, 3:4)
```
```{r gr3}
# make bar graph showing difference for these significant genes 
ecrprisiggr <- ggplot(data = ecrprisig, aes(x = Symbol, y = exp, fill = gr))
ecrprisiggr <- ecrprisiggr + scale_fill_brewer(palette = "Set1")
ecrprisiggr <- ecrprisiggr + geom_bar(stat = "identity", position = "dodge")
ecrprisiggr <- ecrprisiggr + xlab("Genes with Signficant Difference in Groups (p < 0.05)") + ylab("Fold Change")
ecrprisiggr <- ecrprisiggr + ggtitle("Difference in Expression between\nElite Controllers and Rapid Progressors - Recent Infection")
ecrprisiggr <- ecrprisiggr + labs(fill = "Group")
ecrprisiggr <- ecrprisiggr + theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
ecrprisiggr
```

### Make graph into a function

```{r grfunc}
diffgraph <- function(data, titlel2) {
  title <- paste0("Difference in Expression between\n", titlel2)
  ecrpgr <- ggplot(data = data, aes(x = Symbol, y = exp, fill = gr))
  ecrpgr <- ecrpgr + scale_fill_brewer(palette = "Set1")
  ecrpgr <- ecrpgr + geom_bar(stat = "identity", position = "dodge")
  ecrpgr <- ecrpgr + xlab("Genes with Signficant Difference in Groups (p < 0.05)") 
  ecrpgr <- ecrpgr + ylab("Fold Change")
  ecrpgr <- ecrpgr + ggtitle(title)
  ecrpgr <- ecrpgr + labs(fill = "Group")
  ecrpgr <- ecrpgr + theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
  ecrpgr

}
diffgraph(ecrprisig, "Elite Controllers and Rapid Progressors - Recent Infection")
```

## Determine Genes that Are Significantly Different between EC and RP
### 12 Months
### DeltaDeltaCT (2^-DDCT)

```{r modecrp12ddct}
# read file from Excel
ecrp12ddct <- read_csv("ecrp12ddct.csv")
genes <- ecrp12ddct$Symbol
patients <- colnames(ecrp12ddct[2:11])

# set up data frame of t-tests
tt12df <- data_frame(Symbol = genes, tstat = 0.0, pval = 0.0, lowerci = 0.0,
                   upperci = 0.0, sd1 = 0.0, sd2 = 0.0)

# transform matrix to put genes as variables and patients as observations
x <- as_data_frame(t(as.matrix(ecrp12ddct[,2:11])))
colnames(x) <- genes

# set up to loop t-tests over genes
for (i in 1:length(genes)) {
  # set up data frame for this gene from transformed matrix
  # columns 1:5 are group 1 and cols 6:10 are group 2
  # get group 1 data
  ec <- x %>% slice(1:5) %>% select(i)
  ec <- unlist(ec)
  
  # get group 2 data
  rp <- x %>% slice(6:10) %>% select(i)
  rp <- unlist(rp)
  
  # Calculate standard deviations for both groups and place in data frame
  tt12df$sd1[i] <- sd(ec, na.rm = TRUE)
  tt12df$sd2[i] <- sd(rp, na.rm = TRUE)
  
  # run t.test and load values in data frame
  xt <- t.test(ec, rp)
  tt12df$tstat[i] <- xt$statistic
  tt12df$pval[i] <- xt$p.value
  tt12df$lowerci[i] <- xt$conf.int[1]
  tt12df$upperci[i] <- xt$conf.int[2]
}

```

```{r modsiggenes12}
siggenes12 <- tt12df %>% 
            filter(pval < 0.05) %>%
            arrange(pval)
kable(siggenes12)
# Create data frame with expression data for these significant genes
genessig12 <- siggenes12$Symbol
ecrp12sig <- mzdata %>% filter(Symbol %in% genessig12) %>%
             select(Position, Symbol, ECRI, RPRI, diffecrpri) %>%
             gather(key = gr, value = exp, 3:4)
diffgraph(ecrp12sig, "Elite Controllers and Rapid Progressors - 12 months")
```